Equity conceptualizations and standards within transportation literature: a scoping review

### 2023-10-23 Manuscript/Manuscript.qmd
Almost done revision - what is remaining to be tied up is: - "What" are the objects of transportation justice - "How" is fairness determined" - Call 5 - Concluding paragraph

```{r by-year-and-continent-plot}
year_cont_plot <- ggplot(data = Consensus_only_analysis,
                         aes(y = `_Publication year`, 
                         fill = `_Case study continent`)) +
  geom_bar(aes(x = (..count..)/sum(..count..)),col="black") + 
  scale_x_continuous(labels=scales::percent,
                     breaks = seq(0,0.2, by=0.05),
                     limits = c(0,0.2)) +
  xlab("% of papers") +
  theme_classic() + 
  theme(text=element_text(size=10),
        axis.title.y = element_blank(),
        legend.position = "bottom") + guides(fill = guide_legend(nrow = 1,
                                                                 title=NULL))
```

@@

The question of _fairness_, however, is not a simple one, and most of us would probably have been stumped to explain in a clear and precise way, beyond a disquieting intuition, just why the above picture was disturbing. _Justice_ is a political ideal based on the principle that individuals should be treated in a _fair_ and _equitable_ manner [@gosslingUrban2016, p. 2], giving and receiving whatever they are _due_ [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2]. The political [and contested; @vanoutriveHow2019] nature of the concept presents challenges that are only narrowly amenable to scientific inquiry. For starters, the notion that people are "due" something depends on the values of a society, as embodied in its systems and institutions [@karnerFrom2020]. Values, in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe. For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all—perhaps including their own time or bodily autonomy—to some collective illusion (e.g., the state, or a monarch). As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit). In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be what the state, the monarch, or the extremely wealthy said it was. In contrast, in democratic societies^[Paraphrasing, democracy is the worst of collective illusions, except for all others.] individuals rights are valued above the whims of the few. In such a setting, the task of defining a "just" distribution of the burdens and benefits of things—from income, to roads, to space travel—quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.

The rocket that took Bezos to the edge of space is a somewhat rare example of a transportation technology, a tool of space-time convergence. By enabling movement at very high speeds, rockets might—one day, at some indeterminate point in the future—prove essential to the expansion of the human species beyond our home planet. But in the present moment, the benefits of a private suborbital flight (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space) are for a few, whereas the burdens (e.g., the use of non-renewable resources, the climate-altering emissions) affect us all, and not evenly at that. The benefits of public transportation, a much more common transportation technology, are for most, but in many cases we have penny-pinched these systems while concentrating the costs on those who have less [@allenSuburbanization2020; @kaeorueanAnalysis2020]. In a plutocratic society, those with most can (and often do argue^[See the so-called [Techno-optimist Manifesto](https://a16z.com/the-techno-optimist-manifesto/)]) that this distribution of burdens and benefits is fair since both benefits and burdens are earned. The fact that the likes of Bezos travel to space is proof that the likes of Bezos are due those trips. In a democratic society, the members of the collective might actually agree that large rewards (e.g., space flight) must be offered to highly qualified individuals (e.g., Bezos) to entice them to take important responsibilities (e.g., founding and leading Amazon). The fact that some must give up jobs because it takes them too long to reach them by public transportation would constitute proof that those people should have studied more, been earlier risers, worked harder [@SpurrWaiting2015; @greismanLittle2017]. Again, the members of the collective might think this a fine state of affairs, having come to this conclusion of their own accord or after being persuaded by billionaires. Or, contrariwise, the members of the collective might decide that this state of affairs is _unjust_: the _values_ of those in charge of defining what is "fair" matter.

Multiple national and cross-national studies suggest that people in many societies do indeed have some tolerance for inequality [@kiatpongsanHow2014]: it would appear that _some_ stratification, as suggested by @davisSome1945, is perceived as fulfilling a valuable function. However, extreme inequality is often frowned upon, and can lead to social dysfunction and other ills [@acemogluWhy2000; @taydasCan2012; @duIncome2019; @houleInequality2022]. But the perceptions of what is "fair" are neither universal or static. Instead, they are malleable, and can be affected by the existence of opportunities for social mobility [@shariffIncome2016; @artigeWhat2019], by exposure to inequality [@schroderIncome2017; @garciaChanging2023], by learned helplessness [@kimSocioeconomic2022], and even by how information about inequality is communicated to the public [@walkerPeople2021]. It follows that, in general terms, there are at least three different manners of thinking about inequality^[@martensTransport2016, in the introduction of his landmark text, talks about explanatory and prescriptive theories of justice; despite the saying "the arc of history is long and bends toward justice", it is unlikely that a predictive theory of justice exists.]: 1) in a _positive_ (or descriptive) manner, as the current state of the distribution of benefits and burdens of things; 2) also descriptively, as the _perceptions_ about the distribution of those benefits and burdens; and 3) normatively (or prescriptively), as the desired or ideal state of the same. Clearly, the three may coincide (e.g, if the perceived levels of inequality matched actual inequality and also how much of it the public desired). However, they do not necessarily have to, and in many cases will differ from one another. Measuring inequality as it is, in the most objective way possible is an essential task to decide whether there is a need to develop inequality-related policy to increase fairness. In turn, measuring the perceptions of inequality, in the most accurate way possible, may be important to achieve sufficient public buy-in in order to enhance the chances that a given policy will succeed.

Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation [@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009], have increasingly come into focus from the perspective of equity. In response to this focus, a lively and rapidly growing literature has emerged on the topic [see _inter alia_ @martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023]. A cynical rationale for this interest could be that keeping track of actual and perceived inequalities can serve _at the very least_ as a gauge of social discontent [as Chilean authorities discovered to their sorrow when unrest erupted over the increase of metro fares in 2019; @BBCNews2019; @diazpabonInequality2021]. More optimistically, in democratic systems, there could be a genuine interest in being ready to respond to popular demands for fairness. Several challenges arise when approaching this endeavor, caused by the notorious complexity of transportation systems: simultaneously, they move people, goods, and information. As well, emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020]. Unlike income, which is as unidimensional as it gets, the benefits and burdens of transportation systems are diffuse over space and time. For example, transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK. Furthermore, the shades of policies past can continue to haunt a region and even the planet for decades or longer, as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change to be fully felt [@markolfTransportation2019].

@@

After this introduction, we set the stage for our investigation by laying out some important definitions. We then describe the methods used for searching, selecting, and reviewing the relevant literature. This is followed by a description of the landscape surveyed, which is then appraised critically. We will conclude with some calls for action to improve the practice of setting and using standards of fairnes for transportation justice.

@@

It is said that justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2], and it ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440].  Justice is a fluid concept, because it depends on the desirability of different states of affairs, which may change between populations or over time. That said, it is possible to distinguish several forms of justice, including [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:

@@

The term "equity" tends to encompass various tools to understand the distribution of benefits and burdens of things among a population. In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020]. Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities. In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, but they are also vulnerable to misuse and even abuse, as tools of subjugation.

Fairness, in contrast to equity, is somewhat more complicated to define. The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways. That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning in the US. Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice). What kind of justice does fairness serve in each case? It depends on what was the reason for seeking justice in the first place. Our reading of @martensFair2021 is that fairness is a yardstick that is best deployed _a priori_ than _a posteriori_, for doing the latter risks rationalizing the outcomes rather than driving them.

@@

To quickly recap the discussion so far, justice is an end. But to understand what that end is, we must clearly define standards of fairness. Equity analysis is tool to measure where the actual or perceived distribution of the burdens and benefits of the products of transportation systems stand with respect to the standard, in other words, instruments to see how close or far we are from a just situation.

@@

- **"Where?"** Traditionally related to the applicable domain or sphere of life relevant for justice. Conventionally this meant the in-group [e.g., members of the same nation state; @jaggarPhilosophicalChallengesGlobal2009, p. 3], a notion since expanded (e.g., with the Universal Declaration of Human Rights) to include all members of the human species. In the case of justice in transportation, the question of "where?" is paramount; it might be argued that, by their very nature, transportation generates inequalities. By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated. The burdens of transportation, in contrast, are often diffuse. They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health.

- Conventionally, the question of **"When?"** refers to the situation, that is, the social circumstances within which the demands of justice have application. In our case, we might as well ask about the temporal aspects of transportation systems, which include _when_ to invest in transportation [e.g., @rabelloEvaluation2015] (and as a result when to generate a spatial inequality), for how long the burdens and benefits can still be associated to a specific intervention, and whether claims to reparative justice expire over time. <!-- WHEN IN TIME SOMETHGING HAPPENS TOO - WHAT INVESTMENTS HAPPENED IN THE PAST THAT CAST SHADOWS-->

- When we ask **"Who?"**, we inquire about which entities should be regarded as subjects of justice,<!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'. LIKE INCOME, THE MOBILITY TOOL IS MODIFIABLE -- THE MOBILITY TOOL IS NOT MODIFIABLE IN MANY CASES, IT IS WHO YOU ARE . T IS ALMOST LIKE A CHARACTERISTICS OF THE PERSON. PART OF THE SOCIAL CONTRACT THAT IDENTIFIES EOPEL WITH THEIR MOBILITY TOOL--> meaning those entitled to make claims of moral consideration from the perspective of justice. To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income. Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different between each other, in addition to being different from those of men. A possible complication in the case of transportation is that disentangling the "who" from the tools available to them is not always straightforward. Clearly, a person is not their mode of transportation; in practice, though, there are large segments of the population who live in situations where they cannot extricate themselves from the tools for mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control, effectively ending up as captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]. In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume. So, while it is important to avoid conflating the "who" with the "what", for analytical purposes we need to be mindful of the connection between person and their mobility tools. In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on. These could be ministries or departments of transportation, transit agencies, public works departments and others. Identifying these entities is relevant to ellucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.

- **"What?"** asks which entities should be regarded as objects of justice, meaning which kinds or categories of things should be distributed in a just manner. To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. In other words, what do transportation systems _produce_? At their most fundamental, transportation systems are space-time convergence technologies, tools that improve the rate at which time is traded for space. They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to take off. With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement. The realization of this potentials happens through travel. However, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013]. For this reason, we cannot stop at considering mobility, but the ulterior goal of mobility, which is to reach destinations. In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach destinations. In this manner, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation). The burdens of transportation are also many and varied. Some are direct and paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution).

- The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances ("when" and "where"). **Equity standards** are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom? Standards are thresholds that when operationalized effectively define what is fair, that is, in/equitable. The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two. Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel. A number of theoretical and conceptual frameworks exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, and sufficientarism.

>>>>

```{r by-year-and-continent-plot}
year_cont_plot <- ggplot(data = Consensus_only_analysis,
                         aes(y = `_Publication year`, 
                         fill = `_Case study continent`)) +
  geom_bar(aes(x = (..count..)/sum(..count..)),col="black") + 
  scale_x_continuous(labels=scales::percent,
                     breaks = seq(0,0.2, by=0.05),
                     limits = c(0,0.2)) +
  xlab("% of papers") +
  theme_classic() + 
  theme(text=element_text(size=10),
        axis.title.y = element_blank(),
        legend.position = "bottom") + guides(fill = guide_legend(nrow = 2,
                                                                 title=NULL))
```

@@

The question of _fairness_ is not a simple one. Most of us would probably have been stumped to explain it in a <!-- clear and-->precise way<!--, beyond a disquieting intuition,--> just why the above picture was disturbing. _Justice_ is a political ideal based on the principle that individuals should be treated in a _fair_ and _equitable_ manner [@gosslingUrban2016, p. 2], giving and receiving whatever they are _due_ [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2]. The political (and contested; @vanoutriveHow2019) nature of the concept presents challenges that are only narrowly amenable to scientific inquiry. For starters, the notion that people are "due" something depends on the values of a society, as embodied in its systems and institutions [@karnerFrom2020]. Values, in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe. For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all<!-- —perhaps including their own time or bodily autonomy—--> to some collective illusion (e.g., the state, or a monarch). As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit). In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be what organizational structures with power (e.g., the state, the monarch, or the extremely wealthy) said it was. In contrast, in democratic societies^[Paraphrasing, democracy is the worst of collective illusions, except for all others.] individuals rights and a collective vision are valued above the whims of the few. In such a setting, the task of defining a "just" distribution of the burdens and benefits of things —from income, to roads, to space travel- quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.

The rocket that took Bezos to the edge of space is a somewhat rare example of a transportation technology, a tool of space-time convergence. By enabling movement at very high speeds, rockets might—one day, at some indeterminate point in the future—prove essential to the expansion of the human species beyond our home planet. But in the present moment, the benefits of a private suborbital flight (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space) are for a few, whereas the burdens (e.g., the use of non-renewable resources, the climate-altering emissions) affect us all, and not evenly at that. The benefits of public transportation, a much more common transportation technology, are for most, but in many cases we have penny-pinched these systems while concentrating the costs on those who have less [@allenSuburbanization2020; @kaeorueanAnalysis2020]. In a plutocratic society, those with most can (and often do argue^[See the so-called [Techno-optimist Manifesto](https://a16z.com/the-techno-optimist-manifesto/)]) that this distribution of burdens and benefits is fair since both benefits and burdens are earned. The fact that the likes of Bezos do travel to space is proof that the likes of Bezos are due those trips. In a democratic society, the members of the collective might actually agree that large rewards (e.g., space flight) must be offered to highly qualified individuals (e.g., Bezos) to entice them to take important responsibilities (e.g., founding and leading Amazon). The fact that some must give up jobs because it takes them too long to reach them by public transportation would constitute proof that those people should have studied more, been earlier risers, worked harder [@SpurrWaiting2015; @greismanLittle2017]. Again, the members of the collective might think this a fine state of affairs, having come to this conclusion of their own accord or after being persuaded by billionaires. Or, contrariwise, the members of the collective might decide that this state of affairs is _unjust_: the _values_ of those in charge of defining what is "fair" matter.

Multiple national and cross-national studies suggest that people in many societies do indeed have some tolerance for inequality [@kiatpongsanHow2014]: it would appear that _some_ stratification, as suggested by @davisSome1945, is perceived as fulfilling a valuable function. However, extreme inequality is often frowned upon, and can lead to social dysfunction and other ills [@acemogluWhy2000; @taydasCan2012; @duIncome2019; @houleInequality2022]. But the perceptions of what is "fair" are neither universal or static. Instead, they are malleable, and can be affected by the existence of opportunities for social mobility [@shariffIncome2016; @artigeWhat2019], by exposure to inequality [@schroderIncome2017; @garciaChanging2023], by learned helplessness [@kimSocioeconomic2022], and even by how information about inequality is communicated to the public [@walkerPeople2021]. It follows that, in general terms, there are at least three different manners of thinking about inequality^[@martensTransport2016, in the introduction of his landmark text, talks about explanatory and prescriptive theories of justice; despite the saying "the arc of history is long and bends toward justice", it is unlikely that a predictive theory of justice exists.]: 1) in a _positive_ (or descriptive) manner, as the current or historic state of the distribution of benefits and burdens of things; 2) also descriptively, as the _perceptions_ about the distribution of those benefits and burdens; and 3) normatively (or prescriptively), as the desired or ideal state of the same. Clearly, the three may coincide (e.g, if the perceived levels of inequality matched actual inequality and also how much of it the public desired). However, they do not necessarily have to, and in many cases will differ from one another. Measuring inequality as it is and in the most objective way possible, is an essential task to decide whether there is a need to develop inequality-related policies to increase fairness. In turn, measuring the perceptions of inequality, in the most accurate way possible, may be important to achieve sufficient public buy-in in order to enhance the chances that given policies will succeed.

Transportation systems, as a class of essential technologies that facilitate or impede social inclusion and activity participation [@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009], have increasingly come into focus from the perspective of equity. In response to this focus, a lively and rapidly growing literature has emerged on the topic (see _inter alia_ [@martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023]). A cynical rationale for this interest could be that keeping track of objective and perceived inequalities can serve _at the very least_ as a gauge of social discontent (as Chilean authorities discovered when the increase in Santiago Metro's subway fare sparked a period of massive demonstrations and unrest in October 2019 motivated by country-wide social inequalities; [@BBCNews2019; @diazpabonInequality2021]). More optimistically, in democratic systems, tracking objective and perceived inequalities could be of interest for governing bodies to respond to popular demands for fairness. Several challenges arise when approaching this endeavor. The notorious complexity of transportation systems is one challenge: transport systems simultaneously move people, goods, and information, and emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020]. The benefits and burdens of transportation systems are diffuse over space and time. For example, transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off from essential opportunities _elsewhere_, as @rajeEngineering2004 poignantly illustrated with examples of infrastructure in the UK. Furthermore, the shades of policies past can continue to haunt a region and even the planet for decades or longer, as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change to be fully felt [@markolfTransportation2019].

@@

After this introduction, we set the stage for our investigation by laying out some important definitions. We then describe the methods used for searching, selecting, and reviewing the relevant literature. This is followed by a description of the findings from the review, which is then appraised critically. We conclude with some calls for action to improve the practice of setting and using standards of fairness for transportation justice.

@@

It is said that justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2], and it ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440].  Justice is a fluid concept, because it depends on the desirability of different states of affairs, which may change between populations and over time. That said, it is possible to distinguish several forms of justice, including [see @jaggarPhilosophicalChallengesGlobal2009; @pereiraDistributiveJusticeEquity2017a; @karnerTransportationEquityTransportation2020]:

@@

The term "equity" tends to encompass various tools to understand the distribution of benefits and burdens of things among a population, with a particular emphasis on those with the least advantage and equality in their outcomes. In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020]. Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities. In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, or else they may be more vulnerable to misuse and even abuse as tools of subjugation.

Fairness, in contrast to equity, is somewhat more complicated to define. The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways. That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning in the US. Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice). What kind of justice does fairness serve in each case? It depends on what was the rationale for seeking justice in the first place. Our reading of @martensFair2021 is that fairness is a yardstick that is best deployed _a priori_ than _a posteriori_, for doing the latter risks rationalizing the outcomes rather than driving them.

@@

To recap the discussion of definitions so far, justice is an end. But to understand what that end is, we must clearly define standards of fairness. Equity analysis is tool to measure where the actual or perceived distribution of the burdens and benefits of the products of transportation systems stand with respect to the standard, in other words, instruments to see how close or far we are from a just situation.

@@

- **"Where?"**: Such questions traditionally relate to the applicable domain or sphere of life relevant for justice. Conventionally this meant the in-group e.g., members of the same nation state [@jaggarPhilosophicalChallengesGlobal2009, p. 3]<!--, a notion that can be expanded to include all members of the human species (e.g., with the Universal Declaration of Human Rights) and nature (e.g., such as the Ecuadorian constitution, the first ever country to grant legal rights to nature [@akchurinConstructingRightsNature2015])-->. In the case of justice in transportation, the question of "where?" is paramount; as it might be argued that, by their very nature, transportation generates inequalities. By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated. The burdens of transportation, in contrast, are often diffuse. They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health. As such, the definition of the spatial boundaries in the analysis is the answer to "where?".

- Conventionally, the question of **"When?"** refers to the temporal circumstances within which the demands of justice have application. In the cause of transportation justice, we ask about the temporal aspects of transportation systems, as examples: _when_ did the equity analysis take place and under what historical policy context, _the right time_ to invest in transportation infrastructure [e.g., @rabelloEvaluation2015] (and as a result when to generate a spatial inequality), for _how long_ the burdens and benefits can still be associated to a specific transportation intervention, or even _timelines_ of reparative justice interventions that reconcile the shadows of past transportation-related injustices.

- When answering **"Who?"**, we inquire about which entities should be regarded as subjects or arbiters of justice,<!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'. LIKE INCOME, THE MOBILITY TOOL IS MODIFIABLE -- THE MOBILITY TOOL IS NOT MODIFIABLE IN MANY CASES, IT IS WHO YOU ARE - LIKE A CHARACTERISTICS OF THE PERSON. PART OF THE SOCIAL CONTRACT THAT IDENTIFIES PEOPLE WITH THEIR MOBILITY TOOL--> meaning those entitled to make claims of moral consideration from the perspective of justice. To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income. Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different between each other, in addition to being different from those of White men. A possible complication in the case of transportation is that disentangling the "who" from their mobility tools is not always straightforward. Clearly, a person is not their mode of transportation; however in practice, there are large segments of the population who live in situations where they cannot extricate themselves from the tools for mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control (e.g., captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]). In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume. So, while it is important to avoid conflating the "who" with the "what", for analytical purposes we need to be mindful of the connection between person and their mobility tools. In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on. These could be ministries or departments of transportation, transit agencies, public works departments and others having the power to act upon transportation (in)justices. Identifying these entities is relevant to ellucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.

- **"What?"** asks which entities should be regarded as objects of inequities, meaning which kinds or categories of things should be distributed in a just manner. To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. In other words, what do transportation systems _produce_? At their most fundamental, transportation systems are space-time convergence technologies, tools that improve the rate at which time is traded for space. They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to take off. With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement. The realization of this potentials happens through travel. However, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013]. For this reason, we cannot stop at considering mobility, but the ulterior goal of mobility, which is to reach destinations. In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach destinations. In this manner, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation). The burdens of transportation are also many and varied. Some are direct and paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution).

- The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances ("when" and "where"). **Equity standards** are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom? Standards are thresholds that when operationalized effectively define what is fair, that is, (in)equitable. The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two. Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel. A number of theoretical and conceptual frameworks exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, and sufficientarism.

- Lastly, convincing answers to the above questions require a supporting rationale: a **"Why?"** [@jaggarPhilosophicalChallengesGlobal2009]. This is perhaps the most slippery of all the questions posed here. Justice is an inherently social construct<!--, and does not exist outside of human society-->. Asking **why?** amounts to asking what sort of social contracts regulate human interactions, that is, what are the rules that our collective will to believe imposes on each of us. These contracts can be defined by constitution, but there are often unwritten and possibly contested variants. To give an example drawing from the Canadian Charter of Rights and Freedoms, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex". Notice that this declaration includes several individual traits that define the subjects of justice, but does not consider age as one aspect of the person. Does this mean that they apply universally irrespective of age? Surely not, since no reasonable person would consider a toddler's demands for freedom of association as an absolute. So, when do these rights fully apply in the life of a person? In 2017, Adrian Crook of Vancouver, B.C., was warned by the province's Ministry of Children and Family Development that his kids could not be out of home in the community, alone or with other kids the same age, without supervision. The Ministry's argument was that children riding the bus unsupervised compromised the parent's ability to provide care and placed them at risk [@brendHere2017]. Crook argued that the goal in teaching his kids (then aged 7, 8, 9 and 11) to use transit to go to school was to raise self-reliant children [@stueckYoung2017]. While both sides discovered that no clear rules existed regarding childrens' independent transit use [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018]. If the public good is to reduce risk, shouldn't children be banned from riding in cars? The Canadian legal system concerns itself with fundamental freedoms and rights, one of which is the liberty to move freely. This provides an answer (of many) as to "why" we would even consider children as subjects of justice. But the devil is in the details, and answering the other questions discussed above is essential to pin the devil by the tail. Are children under the age of 10 subjects of justice (i.e., are they the "who") and who is deciding if they are (i.e., the courts and the parents in this case)? If so, is the ability to use transit unsupervised an object of justice that should be justly distributed (i.e., the "what") and under what circumstances (e.g, to school, to a social event; during the day, or late at night?). "When" (e.g., for all children going forward, for only 5 years) and "where" (e.g., only in the suburbs, across the country) do these questions apply? Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire independent mobility [@stueckCourt2020]. In actuality, the courts _created_ a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.

Another perspective on the answer to "Why?" are cultural _norms_ in providing rationale. In this case, the liberty of movement for all - including children - reigns supreme. But reigns supreme, "Why"? The outside is dangerous, in part due to vehicular traffic [@morencyTraveling2018], and in part because of potential violence from others. Parents may have over-pronounced 'stranger danger' based on neighbourhood and socio-economic characteristics of neighbourhoods [@francisLlBeDriving2017] and children themselves may fear potential violence based on real threats of victimization [@painParanoidParentingRematerializing2006]. Who's voice, what perception of fear, and actions we take to intervene define and reproduce cultural norms, all play into the rationale behind the question "Why?". In this way, analyzing the "Why?" in the reviewed literature is left out of this review as answers to "Why?" are not explicitly stated and challenging to infer implicitly. As such, the focus of this report is on the standards of fairness that, combined with the use of equity analysis, can help us understand how better to move towards just transportation systems and better formulate answers to "Why?".
